Categorical Foundations of E-Graphs With Binding

### 2025-07-11 POPL2026/paper/CategoricalFoundationsOfEGraphsWithBinding.tex
save the trees
cosmetic changes

\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{stmaryrd}

@@

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{lemma}[theorem]{Lemma}

@@

\begin{abstract}
	Equality saturation, a technique for program optimisation and reasoning, has gained attention due to the resurgence of equality graphs (e-graphs).
	E-graphs represent equivalence classes of terms under rewrite rules, enabling simultaneous rewriting across a family of terms.
	However, they struggle in domains like $\lambda$-calculus that involve variable binding, due to a lack of native support for bindings.
	Building on recent work interpreting e-graphs categorically as morphisms in symmetric monoidal semilattice-enriched categories, we extend this framework to closed symmetric monoidal semilattice-enriched categories to handle bindings.
	We provide a concrete combinatorial representation using an extension of hierarchical hypergraphs and introduce a corresponding double-pushout (DPO) rewriting mechanism.
	Finally, we establish the equivalence of term rewriting and DPO rewriting, with the key property that the combinatorial representation absorbs the equations of the symmetric monoidal structure.
\end{abstract}

>>>>

\usepackage{amsthm}
\usepackage{hyperref}
\renewcommand{\sectionautorefname}{Section}
\renewcommand{\subsectionautorefname}{Subsection}

\usepackage{cleveref}
\usepackage{stmaryrd}

@@

\usepackage{aliascnt}

% space saving
\usepackage[moderate,bibbreaks=normal,bibnotes=normal,bibliography=normal,leading=normal]{savetrees}

@@

\newtheorem{theorem}{Theorem}[section]

% definition
\newaliascnt{definition}{theorem}
\newtheorem{definition}[definition]{Definition}
\aliascntresetthe{definition}
\providecommand*{\definitionautorefname}{Definition}

\newaliascnt{example}{theorem}
\newtheorem{example}[example]{Example}
\aliascntresetthe{example}
\providecommand*{\exampleautorefname}{Example}

\newaliascnt{proposition}{theorem}
\newtheorem{proposition}[proposition]{Proposition}
\aliascntresetthe{proposition}
\providecommand*{\propositionautorefname}{Proposition}

\newaliascnt{remark}{theorem}
\newtheorem{remark}[remark]{Remark}
\aliascntresetthe{remark}
\providecommand*{\remarkautorefname}{Remark}

\newaliascnt{lemma}{theorem}
\newtheorem{lemma}[lemma]{Lemma}
\aliascntresetthe{lemma}
\providecommand*{\lemmaautorefname}{Lemma}

@@

\begin{abstract}
	% Equality saturation, a technique for program optimisation and reasoning, has gained attention due to the resurgence of equality graphs (e-graphs).
	% E-graphs represent equivalence classes of terms under rewrite rules, enabling simultaneous rewriting across a family of terms.
	% However, they struggle in domains like $\lambda$-calculus that involve variable binding, due to a lack of native support for bindings.
	% Building on recent work interpreting e-graphs categorically as morphisms in symmetric monoidal semilattice-enriched categories, we extend this framework to closed symmetric monoidal semilattice-enriched categories to handle bindings.
	% We provide a concrete combinatorial representation using an extension of hierarchical hypergraphs and introduce a corresponding double-pushout (DPO) rewriting mechanism.
	% Finally, we establish the equivalence of term rewriting and DPO rewriting, with the key property that the combinatorial representation absorbs the equations of the symmetric monoidal structure.
    Equality saturation is a powerful technique for program optimisation and reasoning, driven by the use of equivalence classes of terms under rewrite rules called equality graphs (e-graphs). 
    Despite their numerous advantages, until recently e-graphs lacked native support for variable binding, limiting their applicability to programming languages. 
    We propose to address this problem from a categorical perspective, interpreting e-graphs as string diagrams, namely morphisms in symmetric monoidal semilattice-enriched categories with a monoidal closed structure. 
    We further define a concrete representation using extended hierarchical hypergraphs, and introduce a corresponding double-pushout (DPO) rewriting system. 
    Finally, we establish the equivalence between term rewriting and DPO rewriting, with the combinatorial model inherently absorbing the equations of the symmetric monoidal structure. 
    Our approach, specifically designed for lambda calculi, is compared and contrasted with slotted e-graphs—a recent, more generic method for incorporating variable binding.
\end{abstract}

### 2025-07-11 POPL2026/paper/sections/appendix.tex
cosmetic changes

@@

In constructing the pushout the functional versions of e-hypergraph relations will be useful.
\begin{remark}
	$\textcolor{e-color}{<}$, $\textcolor{closed-color}{<}$ and $\consistency$ can be considered as (partial) functions defined on $V_{\mathcal{F}} + E_{\mathcal{F}}$, \textit{i.e.}, on the coproduct of vertices and edges.
	To make things well-typed, we will use corresponding coproduct injections $\iota_{V} : {V_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ and $\iota_{E} : {E_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ when passing either a vertex or an edge into these functions.
	For example, an immediate successor of a vertex $x$ can be written functionally as $\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))$.
\end{remark}

@@

		\item When $x$ is not a top-level vertex such that $\textcolor{e-color}{<}(\iota_{V}(x))$ is defined,
		      \[
			      \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
		      \]
		      and
		      \[
			      \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \] when $x$ is a not top-level edge.
		\item When $x$ is not a top-level vertex such that $\textcolor{closed-color}{<}(\iota_{V}(x))$ is defined,
		      \[
			      \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
		      \]
		      and
		      \[
			      \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \] when $x$ is a not top-level edge.
		\item
		      When $x \in E_{\mathcal{F}}$
		      \[
			      [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
			      \subseteq
			      \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \]
		      where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}] : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
		\item When $x \in V_{\mathcal{F}}$
		      \[
			      [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
			      \subseteq
			      \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x)).
		      \]
	\end{enumerate}
\end{definition}

\begin{theorem}[Existence of pushouts in $\catname{EHyp}(\Sigma)$]
	\label{th:existence_of_pushouts}
	Consider the following span in $\catname{EHyp}(\Sigma)$
	% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
	\[\begin{tikzcd}
			Z && X \\
			\\
			Y
			\arrow["f", from=1-1, to=1-3]
			\arrow["g"', from=1-1, to=3-1]
		\end{tikzcd}\]
	such that
	\begin{enumerate}
		\label{pushout:assumptions}
		\item $Z$ is a \textit{discrete} e-hypergraph.
		\item \label{assumption:equal_predecessors} $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
		\item \label{assumption:non_ambiguous_predecessors} If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
		\item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
	\end{enumerate}
	then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}

	We next explicitly construct a pushout.
	Consider the diagram below.
	\[
		\adjustbox{scale=1.25}{
			\begin{tikzcd}
				Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
				Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
				&                                                                  & Q
			\end{tikzcd}}
	\]
	Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
	First, a coproduct of $X+Y$ is computed which is
	\[
		X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
	\]
	where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
	We will omit labels as they are irrelevant to pushout construction.
	Consider relations
	\begin{align*}
		S_{V} & = \{
		\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}
		\}                                                                                                       \\
		\\
		S_{E} & = \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
	\end{align*}
	and let relations $R_{V},R_{E}$ be their transitive closures respectively.

	We then quotient the set vertices and edges in $X + Y$ by $R_{V}$ and $R_{E}$ respectively.

	\begin{align*}
		\sfrac{X + Y}{\sim (R_{V},R_{E})} = & (
		\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},                                                             \\
		                                    & \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},                       \\
		                                    & \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},                  \\
		                                    & \textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}       \\
		                                    & \textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}})
	\end{align*}

	We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and concrete relation will be clear from the context.
	Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
	We have
	\[
		\sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
	\]
	and there is an obvious surjective function $[]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[]_{V}^{*}$ is its extension to sequences.
	There is also $[] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
	We then define
	\[
		\sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[]^{*}(e) = [s_{X+Y}(e)]^{*}
	\]
	We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
	Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
	Likewise,
	\[
		t_{\sfrac{X+Y}{\sim}}([e]) = [t_{X+Y}(e)]^{*}
	\]

@@

Consider an example of EDPOI rewriting in~\autoref{fig:dpoi-example} as defined in~\autoref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
	[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of~\autoref{def:dpoi-e}.

>>>>

\section{Hypergraphs}
\label{sec:appendix:hyp}
\autoref{fig:hypergraph-with-interfaces} gives an example of an MDA cospan, which is such because it \enquote{looks like} the string diagram \comonoid.
Analogously, we can see that~\autoref{fig:hypergraph} (equipped with the interfaces choosing $\{ v_1 \}$ and $\{ v_4, v_5, v_6, v_7 \}$) fails to form an MDA cospan --- it is not monogamous for many reasons: $v_1$ has out-degree 0, $v_1, v_5, v_7$ have in-degree 2, and $v_3$ has in-degree 0; additionally, it fails be be directed acyclic due to the path along $e_4$ from $v_7$ to itself\footnote{Recall that loops are not part of the syntax of string diagrams for symmetric monoidal categories in the absence of a trace --- a relaxation of monogamy to allow for this is explored by~\citet{ghica_rewriting_2023}.}.

\begin{figure}
	\begin{subfigure}[c]{0.3\linewidth}
		\centering
		% \begin{tikzpicture}[
		% 		vertex/.style={draw, circle, fill, inner sep=1.5pt},
		% 		hyperedge/.style={draw, rectangle}
		% 	]
		% 	\node[vertex, label=left:{$v_4$}] (v1) at (0,0) {};
		% 	\node[vertex, label=left:{$v_5$}] (v2) at (1,0) {};
		% 	\node[vertex, label=left:{$v_6$}] (v3) at (2,0) {};
		% 	\node[vertex, label=left:{$v_7$}] (v4) at (3,0) {};
		% 	\node[vertex, label=left:{$v_2$}] (v5) at (0.5,-2) {};
		% 	\node[vertex, label=left:{$v_3$}] (v6) at (2,-2) {};
		% 	\node[vertex, label=left:{$v_1$}] (v7) at (2,-3) {};

		% 	\node[hyperedge] (e1) at (0.5,-1) {$e_3$};
		% 	\node[hyperedge] (e2) at (2,-1) {$e_4$};
		% 	\node[hyperedge] (e3) at (0.5,-3) {$e_1$};
		% 	\node[hyperedge] (e4) at (2,-2.5) {$e_2$};

		% 	\draw (v1) -- (e1);
		% 	\draw (v2) -- (e1);
		% 	\draw (v5) -- (e1);
		% 	\draw (v5) edge[bend right] (e3);
		% 	\draw (v5) edge[bend left] (e3);
		% 	\draw (v2) edge[in=130,out=-90] (e2);
		% 	\draw (v3) -- (e2);
		% 	\draw (v4) edge[in=50,out=-90] (e2);
		% 	\draw (v4) edge[in=-50,out=-90] (e2);
		% 	\draw (v6) -- (e2);
		% \end{tikzpicture}
        \[
        \adjustbox{scale=0.8}{
        \tikzfig{figures/hypergraph_example}
        }
        \]
		\caption{Hypergraph with 7 vertices and 4 hyperedges, all directed bottom-to-top; source/target vertices are ordered left-to-right.}%
		\label{fig:hypergraph}
	\end{subfigure}
	\begin{subfigure}[c]{0.3\linewidth}
		\centering
		% \begin{tikzpicture}[
		% 		vertex/.style={draw, circle, fill, inner sep=1.5pt},
		% 		hyperedge/.style={draw, rectangle}
		% 	]
		% 	\node[vertex, label=left:{$v_3$}] (v1) at (0,0) {};
		% 	\node[vertex, label=left:{$v_3$}] (v1') at (0,1.5) {};
		% 	\node[vertex, label=left:{$v_2$}] (v2) at (1,0) {};
		% 	\node[vertex, label=left:{$v_2$}] (v2') at (1,1.5) {};
		% 	\node[vertex, label=left:{$v_1$}] (v3) at (0.5,-2) {};
		% 	\node[vertex, label=left:{$v_1$}] (v3') at (0.5,-3.5) {};

		% 	\node[hyperedge] (e1) at (0.5,-1) {$e_1$};

		% 	\draw (v1) -- (e1);
		% 	\draw (v2) -- (e1);
		% 	\draw (v3) -- (e1);

		% 	\node[draw, rounded corners, fit={(v1'.west) (v2'.west)}, inner sep=12pt] (S) {};
		% 	\node[draw, rounded corners, fit={(v3'.west)}, inner sep=12pt] (T) {};
		% 	\node[draw, rounded corners, fit={(v1.west) (v2.west) (v3.west)}, inner sep=12pt] (H) {};
		% 	\draw[commutative diagrams/hookrightarrow,shorten >=2pt,shorten <=2pt] (S) -- node[left] {$g$} (H);
		% 	\draw[commutative diagrams/hookrightarrow,shorten >=2pt,shorten <=2pt] (T) -- node[left] {$f$} (H);
		% \end{tikzpicture}
        \[
        \adjustbox{scale=0.8}{
        \tikzfig{figures/cospan_hypergraph_example}
        }
        \]
		\caption{Hypergraph with interfaces}%
		\label{fig:hypergraph-with-interfaces}
	\end{subfigure}
	\caption{Example hypergraphs}%
	\label{fig:hypergraph-examples}
\end{figure}

@@

In constructing the pushout the functional versions of e-hypergraph relations will be useful.
\begin{remark}
	$\textcolor{e-color}{<}$, $\textcolor{closed-color}{<}$ and $\consistency$ can be considered as (partial) functions defined on $V_{\mathcal{F}} \sqcup  E_{\mathcal{F}}$, \textit{i.e.}, on the coproduct of vertices and edges.
	To make things well-typed, we will use corresponding coproduct injections $\iota_{V} : {V_{\mathcal{F}}} \to V_{\mathcal{F}} \sqcup  E_{\mathcal{F}}$ and $\iota_{E} : {E_{\mathcal{F}}} \to V_{\mathcal{F}} \sqcup  E_{\mathcal{F}}$ when passing either a vertex or an edge into these functions.
	For example, the immediate predecessor of a vertex $x$ can be written functionally as $\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))$.
\end{remark}

@@

		\item When $x$ is not a top-level vertex such that $\textcolor{e-color}{<}(\iota_{V}(x))$ is defined,
		      \[
			      \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
		      \]
		      and
		      \[
			      \phi_{E}(\textcolor{e-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{e-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \] when $x$ is a not top-level edge.
		\item When $x$ is not a top-level vertex such that $\textcolor{closed-color}{<}(\iota_{V}(x))$ is defined,
		      \[
			      \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{V_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x))
		      \]
		      and
		      \[
			      \phi_{E}(\textcolor{closed-color}{<_{\mathcal{F}}^{\mu}}(\iota_{E_{\mathcal{F}}}(x))) = \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \] when $x$ is a not top-level edge.
		\item
		      When $x \in E_{\mathcal{F}}$
		      \[
			      [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}} ]^{*}(\consistency_{\mathcal{F}}(\iota_{E_{\mathcal{F}}}(x)))
			      \subseteq
			      \consistency_{\mathcal{G}}(\phi_{E};\iota_{E_{\mathcal{G}}}(x))
		      \]
		      where $\phi_{V};\iota_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} \sqcup  E_{\mathcal{G}}$, and similarly for $\phi_{E};\iota_{E_\mathcal{G}}$ so that $[\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}] : V_{\mathcal{F}} \sqcup  E_{\mathcal{F}} \to  V_{\mathcal{G}} \sqcup  E_{\mathcal{G}}$.
		\item When $x \in V_{\mathcal{F}}$
		      \[
			      [\phi_{V};\iota_{V_{\mathcal{G}}}, \phi_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{F}}(\iota_{V_{\mathcal{F}}}(x)))
			      \subseteq
			      \consistency_{\mathcal{G}}(\phi_{V};\iota_{V_{\mathcal{G}}}(x)).
		      \]
	\end{enumerate}
\end{definition}

\begin{theorem}[Existence of pushouts in $\catname{EHyp}(\Sigma)$]
	\label{th:existence_of_pushouts}
	Consider the following span in $\catname{EHyp}(\Sigma)$
	% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
	\[\begin{tikzcd}
			Z && X \\
			\\
			Y
			\arrow["f", from=1-1, to=1-3]
			\arrow["g"', from=1-1, to=3-1]
		\end{tikzcd}\]
	such that
	\begin{enumerate}
		\label{pushout:assumptions}
		\item $Z$ is a \textit{discrete} e-hypergraph.
		\item \label{assumption:equal_predecessors} $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
		\item \label{assumption:non_ambiguous_predecessors} If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
		\item $\consistency(f_{V}(v_i)) = \consistency(f_{V}(v_j))$ and $\consistency(g_{V}(v_i)) = \consistency(g_{V}(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
	\end{enumerate}
	then the pushout $X \sqcup _{f,g} Y$ exists.
\end{theorem}
\begin{proof}

	We next explicitly construct a pushout.
	Consider the diagram below.
	\[
		\adjustbox{scale=1.25}{
			\begin{tikzcd}
				Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
				Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X\sqcup Y}{\sim R} \arrow[rd, "u"]                              &   \\
				&                                                                  & Q
			\end{tikzcd}}
	\]
	Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
	First, a coproduct of $X\sqcup Y$ is computed which is
	\[
		X \sqcup  Y = \{V_{X} \sqcup  V_{Y}, E_{X} \sqcup  E_{Y}, s_{X\sqcup Y}, t_{X\sqcup Y}, \consistency_{X\sqcup Y}, \textcolor{e-color}{<_{X\sqcup Y}}, \textcolor{closed-color}{<_{X\sqcup Y}}\}
	\]
	where $s_{X\sqcup Y} : E_{X} \sqcup  E_{Y} \to (V_{X} \sqcup  V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} \sqcup  E_{Y} \to (V_{X} \sqcup  V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} \sqcup  V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} \sqcup  V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X\sqcup Y}, \textcolor{e-color}{<_{X\sqcup Y}}, \textcolor{closed-color}{<_{X\sqcup Y}}, \consistency_{X\sqcup Y}$.
	We will omit labels as they are irrelevant to pushout construction.
	Consider relations
	\begin{align*}
		S_{V} & = \{
		\;(x_i,y_j) \in (V_{X} \sqcup  V_{Y}) \times (V_{X} \sqcup  V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(y_j,x_j) \in (V_{X} \sqcup  V_{Y}) \times (V_{X} \sqcup  V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(x,x)\;\text{where}\; x \in V_{X} \sqcup  V_{Y}
		\}                                                                                                       \\
		\\
		S_{E} & = \{(x,x) \text{ where } x \in E_{X} \sqcup  E_{Y}\}
	\end{align*}
	and let relations $R_{V},R_{E}$ be their transitive closures respectively.

	We then quotient the set vertices and edges in $X \sqcup  Y$ by $R_{V}$ and $R_{E}$ respectively.

	\begin{align*}
		\sfrac{X \sqcup  Y}{\sim (R_{V},R_{E})} = & (
		\sfrac{V_{X} \sqcup  V_{Y}}{\sim (R_{V},R_{E})},                                                             \\
		                                    & \sfrac{E_{X} \sqcup  E_{Y}}{\sim (R_{V},R_{E})},                       \\
		                                    & \sfrac{s_{X\sqcup Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \consistency_{\sfrac{X\sqcup Y}{\sim (R_{V},R_{E})}},                  \\
		                                    & \textcolor{e-color}{<_{\sfrac{X \sqcup  Y}{\sim (R_{V},R_{E})}}}       \\
		                                    & \textcolor{closed-color}{<_{\sfrac{X \sqcup  Y}{\sim (R_{V},R_{E})}}})
	\end{align*}

	We will refer to $\sim (R_{V},R_{E})$ just as $\sim$, \textit{e.g.}, by writing $\sfrac{V_{X} \sqcup  V_{Y}}{\sim}$ and concrete relation will be clear from the context.
	Where necessary we will refer to $S_{V}$ and $S_{R}$ as $\sim_{S}$.
	We have
	\[
		\sfrac{s_{X\sqcup Y}}{\sim} : \sfrac{E_{X} \sqcup  E_{Y}}{\sim} \to (\sfrac{V_{X} \sqcup  V_{Y}}{\sim})^{*}
	\]
	and there is an obvious surjective function $[]_{V} : (V_{X} \sqcup  V_{Y}) \to (\sfrac{V_{X} \sqcup  V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[]_{V}^{*}$ is its extension to sequences.
	There is also $[] : E_{X} \sqcup  E_{Y} \to \sfrac{E_{X} \sqcup  E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
	We then define
	\[
		\sfrac{s_{X\sqcup Y}}{\sim}([e]) = s_{X\sqcup Y};[]^{*}(e) = [s_{X\sqcup Y}(e)]^{*}
	\]
	We will also use subscripts when it is important to tell if an element of $E_{X} \sqcup  E_{Y}$ has a pre-image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
	Similarly, we will write $v_{x}$ to refer to a vertex with a pre-image in $V_{X}$.
	Likewise,
	\[
		t_{\sfrac{X\sqcup Y}{\sim}}([e]) = [t_{X\sqcup Y}(e)]^{*}
	\]

@@

Consider an example of EDPOI rewriting in~\autoref{fig:dpoi-example} as defined in~\autoref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
	[u_1] \to [u_1, v_1, v_2] \sqcup  [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] \sqcup  [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of~\autoref{def:dpoi-e}.
